---
title: "Example #3: Towards size-based ecosystem models"
author: Julia L. Blanchard
date: Feb, 5, 2020
place: Horbart
output:
  pdf_document: default
  html_document: default
---


# Introduction

In this section of the workshop we will continue exploring how we can learn about models ecological models and data using the "mizer" R package. Here, we will push the boundaries of the theory by introducing species that may not necessarily fit within  the conceptualisaion of the trait-based and size spectrum models. To do so we will think about how different assumptions would need to change to accomodate species beyond fish- such as large baleen whales, seabirds and other species.

To begin with we will use the scaling model - which assumes eggs sizes are notthe same and might be a good representation of a background zooplankton + fish community.

## Part A - The scaling model

```{r}
#get required packages
library(devtools)
#most up to date master branch of mizer
install_github("sizespectrum/mizer")
#documentation here:
#https://sizespectrum.org/mizer/dev/index.html
library(mizer)
require(tidyverse)
options("mizer_new" = TRUE)
```


Let's read in some biomasses of predators in the Southern Ocean.

```{r}

#read in data for prydz bay
dat<-read.csv("data/pred_group_biomasses.csv",header=T)

#read groups for southern ocean model
groups<-readRDS("data/groups.RDS")

groups<-merge(groups,dat,by.x="species",by.y="group",all=T)

#have a look at plot
plot(groups$w_inf,groups$biomass.tperkm2,xlab="log Maximum Weight [g]", ylab=" log Total Biomass", log="xy",col="blue",pch=16)
text(groups$w_inf,groups$biomass.tperkm2,labels=groups$species,cex=0.5)
```

The first question we will pose here is: How do we fit a simple community size spectrum model to community size spectrum data? To answer this we ask: What do the community size spectrum model parameters need to be to best capture observations of the community size spectrum? We will consider only two parameters - the level of fishing effort (here, simply a multiplier of fishing mortality rate) and the background resource carrying capacity which are known to affect the size spectrum.

First let's set up the scaling model and plot it.

```{r}
params <- newTraitParams(max_w_inf = 1e+09)
params <- markBackground(params) 
plotSpectra(params)
sim <- project(params, t_max=100)
plotBiomass(sim)

```

Now we see the size distributions of the background community, we would like to add a species to the foreground. Let's go with a big one! 
```{r}
# set up function to run model and output the predicted size spectrum
species_params <- data.frame(
    species = "baleen.whales",
    w_min = 6e+05, 
    w_inf = 1.03e+08, 
    w_mat = 43250000, 
    beta = 260669.5, 
    sigma = 1.8, 
    z0 = 0,
    alpha = 0.6,
    sel_func = "knife_edge", 
    knife_edge_size = 100, 
    gear = "knife_edge_gear",
    k = 0,
    k_vb = 0.2,
    a = a_m,
    b = b_m
)
params <- addSpecies(params, species_params)
plotSpectra(params)
sim <- project(params, t_max=50)
plotBiomass(sim)

```


We will now use the Rshiny app to explore the effects of parameters values on steady state.

```{r}
tuneParams(params)
```

Clearly, the community isn't providing enough food for baleen whales. Let's have a look at the parameters we've put together for a Southern Ocean ecosystem.

```{r}
#read groups for southern ocean model
groups<-readRDS("data/groups.RDS")
#fix one value 
groups[7,"w_min"]<-4000
# read interaction matrix
# get the interaction matrix
theta <- readRDS("data/prydz_inter.RDS")

mf.params <- newMultispeciesParams(species_params=groups,interaction=theta,n=2/3,p=0.7,q=0.8,f=0.6)
plotSpectra(mf.params)
sim <- project(mf.params, t_max=100)
plotBiomass(sim)
plotSpectra(sim)

```

Let's change some of the parameters which make some of these groups less fishy!
```{r}

####### feeding kernels
params<-mf.params

#new predation kernel varies with pred size and species but is the same for all prey species, need to fill this in with values, according to Ryans' code
pred_kern <- array(0, dim=c(length(params@species_params$species),length(params@w),length(params@w_full)),dimnames=list(params@species_params$species,params@w,params@w_full))

#### change the feeding kernel
pred_kern[]<- mf.params@species_params$beta

mvals<-groups$m
#shift over microzoop
#mvals[1]<--2

for (i in 1:length(mvals)) {
  
  D.z <- 2*(3*params@w*1e12/(4*pi))^(1/3) # convert body mass g to ESD (um)
  betas =  (exp(0.02*log(D.z)^2 - mvals[i] + 1.832))^3
  
  if (!is.na(mvals[i])) pred_kern[i,,]<- betas
  
}
  pred_kern[]<-exp(-0.5*sweep(log(sweep(sweep(pred_kern,3,params@w_full,"*")^-1,2,params@w,"*")),1,params@species_params$sigma,"/")^2)
  pred_kern[] <- sweep(pred_kern,c(2,3),combn(params@w_full,1,function(x,w)x<params@w,w=params@w),"*") # find out the untrues and then multiply
  
####### psi function
  
  # Change allocation to reproduction = check for endotherms/determinate growers (should be 1 at wmat not winf)
 psi_r=5
 params@psi[] <- unlist(tapply(params@w,1:length(params@w),function(wx,Winf,Wmat,n){
   ((1 + (wx/(Wmat))^-psi_r)^-1) * (wx/Winf)^(1-n)},Winf=params@species_params$w_inf,Wmat=params@species_params$w_mat,n=params@n))
# # Set w < 1% of Wmat to 0
 params@psi[unlist(tapply(params@w,1:length(params@w),function(wx,Wmat)wx<(Wmat*0.01),Wmat=params@species_params$w_mat))] <- 0
# # Set all m > M to 1 # Check this is right...
 params@psi[unlist(tapply(params@w,1:length(params@w),function(wx,Winf)(wx/Winf)>1,Winf=params@species_params$w_inf))] <- 1
 
########## erepro
 
 #and erepro - should this decline with Winf or shoudl it just be a very small number 0.005?
#should this actually be closer to 1 for mammals?
params@species_params$erepro <-0.05*params@species_params$w_inf^(-0.75)
# need to think about how this sounds in writing

######### juvenile mortality

## add declining intraspecific juvenile mortality to background mortality 
for (i in 1: length(params@species_params$species)) params@mu_b[i,] <- params@mu_b[i,]  + 0.1*params@w^-0.25



```

Let's take a look.
```{r}
  
params<- setPredKernel(params, pred_kernel = pred_kern)
params@kappa <-100
params <- steady(params)

new.mf<-project(params,t_max=100)
plotBiomass(sim)
plotBiomass(new.mf)

plotSpectra(new.mf)
     
# that went a little funny
  
```

```{r}
tuneParams(params)
```



The next code chunks need to be checked and correcterto do an optimisation

```{r}
# set up function to run model and output the predicted size spectrum

runmodel<-function(parms){# set up and run!
params <- set_community_model(knife_edge_size = 10,kappa=parms$kappa)
sim <- project(params, effort = parms$effort, t_max = 200, dt=0.1)
# normally we would want to pass updated equilibirum values as initial_n and initial_n_pp here, rather than running from initial values each time, but doesn't matter too much for this example
# select last 10 years of output (should be time-averaged)
# slope only
# output<-mean(getCommunitySlope(sim,min_w=16,max_w=1600)[190:200,1])
# or whole size spectrum
output <-apply(sim@n,c(2,3),mean)
return(output)
}

```

```{r}
# set up some initial parameters
parms=data.frame(kappa=0.1,effort=1)

# set up error function to compare predictions with observations (only need range of observations)
sse <- function(parms,dat) {
pred <- log(runmodel(parms)[which(params@w  >= 16 & params@w <= 20000)])
# sum of squared errors, here on log-scale of predictions and data (can change this)
discrep <- pred - dat
return(sum(discrep^2))
}
err<-sse(parms,css$logden)
#test
err

```

```{r}
f <- function (par,dat=css$logden) {
parms$kappa <- par[1]
parms$effort <- par[2]  
sse(parms,dat)
}
# two parameter grid
kappa <- seq(from=0.05,to=0.1,by=0.05)
effort <- seq(from=0,to=2,by=0.1)
grid <- expand.grid(kappa=kappa,effort=effort)
grid$SSE <- apply(grid,1,f)

```


```{r}
# which level of effort has overall least error?
effort.hat <- grid$effort[which.min(grid$SSE)]
kappa.hat <- grid$kappa[which.min(grid$SSE)]
# Basic scatterplot
ggplot(grid, aes(x=effort, y=SSE,col=kappa) ) +
  geom_point() 

```

```{r}
#optimisation
vals<-optim(par=c(0.1,0.1),f,method ="L-BFGS-B",lower=c(1e-3,0),upper =c(1,3))

#plug these ones back into model and compare
params <- set_community_model(knife_edge_size = 10,kappa=vals$par[1])
sim <- project(params, effort = vals$par[2], t_max = 100, dt=0.1)
w <-dimnames(sim@n)$w
n <-sim@n[100,,]

#redo the plot
plot(w,n,typ="l",xlim=c(16,20000),log="xy")
points(exp(css$logw),exp(css$logden), col= "steel blue",cex=0.8,pch=16)


```

