---
title: "Example #3: Towards size-based ecosystem models"
author: Julia L. Blanchard
date: Feb, 5, 2020
place: Horbart
output:
  pdf_document: default
  html_document: default
---


# Introduction

In this section of the workshop we will continue exploring how we can learn about models ecological models and data using the "mizer" R package. Here, we will push the boundaries of the theory by introducing species that may not necessarily fit within  the conceptualisaion of the trait-based and size spectrum models. To do so we will think about how different assumptions would need to change to accomodate species beyond fish- such as large baleen whales, seabirds and other species.

To begin with we will use the scaling model - which assumes eggs sizes are notthe same and might be a good representation of a background zooplankton + fish community.


```{r}
#get required packages
library(devtools)
#most up to date master branch of mizer
#install_github("sizespectrum/mizer")
#install_github("astaaudzi/mizer-rewiring", ref = "temp-model-comp")
#documentation here:
#https://sizespectrum.org/mizer/dev/index.html
library(mizer)
require(tidyverse)

remotes::install_github("sizespectrum/mizerExperimental")
library(mizerExperimental)

```


Let's read in some biomasses of predators in the Southern Ocean.

```{r}

#read in data for prydz bay
dat<-read.csv("data/pred_group_biomasses.csv",header=T)

#read groups for southern ocean model
#groups<-readRDS("data/groups.RDS")
#write.csv(groups,"data/groups.csv")

read.csv("data/groups.csv")
groups<-merge(groups,dat,by.x="species",by.y="group",all=T)

#have a look at plot
plot(groups$w_inf,groups$biomass.tperkm2,xlab="log Maximum Weight [g]", ylab=" log Total Biomass", log="xy",col="blue",pch=16)
text(groups$w_inf,groups$biomass.tperkm2,labels=groups$species,cex=0.5)

# could plot the paramter allometries here to explore
```


## Multispecies (functional group) Southern Ocean Model

Next let's read in the parameters we've put together for a Southern Ocean (Prydz Bay) ecosystem.

```{r}
#read groups for southern ocean model
groups<-read.csv("data/groups2.csv")[-1,]

#groups <- readRDS("data/groups.RDS")[-1,]
#check fails these tests:
groups$w_mat25 >= groups$w_mat 
groups$w_mat25 <= groups$w_min
groups$w_inf <= groups$w_mat25

#[-1,]
#fix one value 
#groups[7,"w_min"]<-4000
# read interaction matrix
# get the interaction matrix
theta <- readRDS("data/prydz_inter.RDS")[-1,-1]
#[-1,-1]

 
```

We will pass these parameters to mizer to set up a new multispecies model.

```{r}
 
mf.params <- newMultispeciesParams(species_params=groups,interaction=theta,kappa = 1e3,min_w_pp = 1e-16,w_pp_cutoff = 1,r_pp = 10,z0pre = 0.01)

# mf.params@species_params$metab[1] <-0.01
mf.params@species_params$h[] <- 10*mf.params@species_params$ks[] 

# mf.params@species_params$gamma[1] <- 1000*(mf.params@species_params$gamma[1])
mf.params@species_params$alpha[] <- 0.6
#mf.params@species_params$z0[1] <- 0.01

#higher mean betas
 mf.params@species_params$beta[mf.params@species_params$species == "large.divers"] <-6000
 mf.params@species_params$beta[mf.params@species_params$species == "apex.predators"] <-1000
# # wider feeding kernels
 mf.params@species_params$sigma[] <- 2
 mf.params@species_params$sigma[mf.params@species_params$species == "baleen.whales"] <- 4


# mf.params@species_params$w_mat25[1] <- 0.005*mf.params@species_params$w_inf[1]
mf.params@species_params$erepro <- ifelse(mf.params@species_params$erepro==1,0.001,mf.params@species_params$erepro)

# mf.params@interaction[] <-0.1
#mf.params@interaction[,1] <-0
#mf.params@interaction[1,1] <-1

#mf.params@species_params$interaction_p[2:11] <-0.5

# hand tuning to gte in line with data
mf.params@species_params$R_max <- 10*mf.params@species_params$R_max


# needs to be very different for marine mammals? use different repro assumptions (fixed offspring density per year)?



# mf.params@species_params$w_mat <- 0.5*mf.params@species_params$w_inf
# mf.params@species_params$w_mat[mf.params@species_params$species == "flying.birds"] <- 0.75*mf.params@species_params$w_inf[mf.params@species_params$species == "flying.birds"]
# mf.params@species_params$w_mat[mf.params@species_params$species == "small.divers"] <- 0.75*mf.params@species_params$w_inf[mf.params@species_params$species == "small.divers"]
# mf.params@species_params$w_mat[mf.params@species_params$species == "medium.divers"] <- 0.75*mf.params@species_params$w_inf[mf.params@species_params$species == "medium.divers"]
# mf.params@species_params$w_mat[10:12] <- 0.75*mf.params@species_params$w_inf[10:12]
# setParams(mf.params)

mf.params<- setParams(mf.params,kappa=1e8)


sim <- project(mf.params, t_max=500,dt = 0.1)
plot(sim)
getFeedingLevel(sim)
#plotlyGrowthCurves(sim,species="macrozooplankton")
plotlyFeedingLevel(sim)
# macrozooplankton - they are not growing enough,why?
#tuneParams(mf.params)
plotlyGrowthCurves(sim,percentage = T)
```

If we combine the biomassess from the model and add the data to  plot we cna look at how close/far we are.

```{r}
# get biomass and covert to tonnes per area
plotBioData <- function (sim,dat){
simbio<-getBiomass(sim)
simbio<-data.frame(species=mf.params@species_params$species,w_inf=mf.params@species_params$w_inf,simbio=simbio[100,]/1e6)

simbio<-as.data.frame(merge(simbio,dat,by.x="species",by.y="group",all=T))

#have a look at plot with both modelled total biomass and observed for top predators
plot(simbio$w_inf,simbio$simbio,xlab="log Maximum Weight [g]", ylab=" log Total Biomass", log="xy",col="grey",
pch=16,ylim=c(min(simbio$simbio,simbio$biomass.tperkm2,na.rm=T),max(simbio$simbio,simbio$biomass.tperkm2,na.rm=T)))
points(simbio$w_inf,simbio$biomass.tperkm2,col="blue",pch=1)
text(simbio$w_inf,simbio$simbio,labels=simbio$species,cex=0.5)
return(simbio)
}

plotBioData(sim,dat)


# 
# 
# ## have a look at paramters - metabolism
# plot(mf.params@species_params$w_inf,mf.params@species_params$ks,log="x")
# text(mf.params@species_params$w_inf,mf.params@species_params$ks,labels=params@species_params$species,cex=0.5)
# 
# ## have a look at paramters - h
# plot(mf.params@species_params$w_inf,mf.params@species_params$h,log="x")
# text(mf.params@species_params$w_inf,mf.params@species_params$h,labels=params@species_params$species,cex=0.5)
# # 

 # plot(mf.params@species_params$w_inf,mf.params@species_params$ks/mf.params@species_params$h,log="x")
 # text(mf.params@species_params$w_inf,mf.params@species_params$ks/mf.params@species_params$h,labels=params@species_params$species,cex=0.5)
# 

# # ## have a look at paramters - gamma
#  plot(mf.params@species_params$w_inf,mf.params@species_params$gamma,log="xy")
#  text(mf.params@species_params$w_inf,mf.params@species_params$gamma,labels=params@species_params$species,cex=0.5)
# # 
# # 
# # ## have a look at paramters - wmat/winf
#  plot(mf.params@species_params$w_inf,mf.params@species_params$w_mat/mf.params@species_params$w_inf,log="xy",xlab=c("Weight"),y=c("Wmat/Winf"))
#  text(mf.params@species_params$w_inf,mf.params@species_params$w_mat/mf.params@species_params$w_inf,labels=params@species_params$species,cex=0.5)
 
# changes need to maturation sizes - zoop groups -> wmat 50% of winf
```

Let's change some of the parameters which make some of these groups less fishy! The feeidng kernel in teh above model is set up for fish. Here we use the model of Heneghan to describe a more planktivours feeding kernel form zooplankton and baleen whales.
```{r}

####### feeding kernels - use box kernels for planktovores instead?
params<-mf.params

#new predation kernel varies with pred size and species but is the same for all prey species, need to fill this in with values, according to Ryans' code
pred_kern <- array(0, dim=c(length(params@species_params$species),length(params@w),length(params@w_full)),dimnames=list(params@species_params$species,params@w,params@w_full))

#### change the feeding kernel
pred_kern[]<- mf.params@species_params$beta

#mvals<-groups$m ### need to enter these manually as removed from file because of new m parameter in mizer setparams function
#shift over microzoop
#mvals[1]<--2

mvals <-c(
NA,
0.73,
NA,
NA,
NA,
NA,
NA,
NA,
NA,
NA,
1.5)


### Ryan's feeding kernel:
for (i in 1:length(mvals)) {
  
  D.z <- 2*(3*params@w*1e12/(4*pi))^(1/3) # convert body mass g to ESD (um)
  betas =  (exp(0.02*log(D.z)^2 - mvals[i] + 1.832))^3
  
  if (!is.na(mvals[i])) pred_kern[i,,]<- betas
  
}
  pred_kern[]<-exp(-0.5*sweep(log(sweep(sweep(pred_kern,3,params@w_full,"*")^-1,2,params@w,"*")),1,params@species_params$sigma,"/")^2)
  pred_kern[] <- sweep(pred_kern,c(2,3),combn(params@w_full,1,function(x,w)x<params@w,w=params@w),"*") # find out the untrues and then multiply
  
##### box feeding kernel for baleen whales  

  
  
  
####### psi function
  
#   # Change allocation to reproduction = check for endotherms/determinate growers (should be 1 at wmat not winf)
#  psi_r=20
#  params@psi[] <- unlist(tapply(params@w,1:length(params@w),function(wx,Winf,Wmat,n){
#    ((1 + (wx/(Wmat))^-psi_r)^-1) * (wx/Winf)^(1-n)},Winf=params@species_params$w_inf,Wmat=params@species_params$w_mat,n=params@n))
# # # Set w < 1% of Wmat to 0
#  params@psi[unlist(tapply(params@w,1:length(params@w),function(wx,Wmat)wx<(Wmat*0.01),Wmat=params@species_params$w_mat))] <- 0
# # # Set all m > M to 1 # Check this is right...
#  params@psi[unlist(tapply(params@w,1:length(params@w),function(wx,Winf)(wx/Winf)>1,Winf=params@species_params$w_inf))] <- 1
#  
# ########## erepro
#  
#  #and erepro - should this decline with Winf or shoudl it just be a very small number 0.005?
# #should this actually be closer to 1 for mammals?
# params@species_params$erepro <-0.05*params@species_params$w_inf^(-0.75)
# # need to think about how this sounds in writing
# 
# ######### juvenile mortality
# 
# ## add declining intraspecific juvenile mortality to background mortality 
# for (i in 1: length(params@species_params$species)) params@mu_b[i,] <- params@mu_b[i,]  + 0.1*params@w^-0.25
# 


```

Let's take a look now.
```{r}
  
params<- setParams(params, pred_kernel = pred_kern,kappa=1e7)


new.mf<-project(params,t_max=100)

plot(new.mf)


new.mf<-project(params,t_max=500,initial_n=new.mf@n[100,,],initial_n_pp=new.mf@n_pp[100,])
   
plot(new.mf)  
plotlyFeedingLevel(new.mf)  
plotlyGrowthCurves(new.mf,percentage = T) 
plotlyBiomass(new.mf,percentage = T) 
plotBioData(new.mf,dat)

```

That helps baleen whales - and impact macrozoooplankton, but some of the predator groups are still not getting enough food.

Next we could use Gustav's RShiny "tuner" to explore the params by  hand to epxlore why this is the case...
```{r}
#tuneParams(params)
```

####### Calibration #######

Rather than manually tune the parameters we will try to automatically calibrate the model with data and estimate our uncertain parameters. We are particularly uncertian about: kappa ( the satellite data used as  inputs does not appear to be enough to keep the zooplankton  grooups alive), Rmax( this will influence the biomass of each group) Here, we will use an inverse calibraion where we have population level biomass density data for 6 of the marine mammal and seabird groups (obtained from the inputs for an EcoPath model). 

```{r}
# Step 0. Set up function to run model and output the predicted log biomasses

runmodel<-function(theta,parms=params,delta_t=0.1){# set up and run!
# change the paramters values that we want to estimate/calibrate
parms@species_params$R_max<- theta[1]*parms@species_params$w_inf^theta[2]
kappa <- theta[3]
backmort <- theta[4]
parms<-setParams(parms,kappa=kappa,z0pre = backmort)
#parms<- setPredKernel(parms, pred_kernel)
sim<- project(parms,dt=delta_t,t_max = 100)
# output the biomass denisty in terms of tonnes per km2
# assume model is set up for an area of 1km2
# this will output the biomass at last time step - do we need a time-average?
sim_logbiomass<-log(getBiomass(sim)[dim(sim@n)[1],]) - log(1e6)
return(sim_logbiomass)
# do we also want to pass the equilibrium n and n_pp from last iteration?
}
#test-run

theta <- c(1e2,-1.5,1e5,0.1)
runmodel(theta = theta,parms=mf.params)
```


```{r}
# set up error function to compare predictions with observations (only need range of observations)
dat<-log(c(rep(NA,4),0.003451, 0.015580,NA,NA,0.258995,0.008246,0.021900,0.141530))
names(dat) <-params@species_params$species

sse <- function(theta, params=params,dat=dat) {
pred <- runmodel(theta,params)
# sum of squared errors, here on log-scale of predictions and data (can change this)
discrep <- pred[!is.na(dat)] - dat[!is.na(dat)]
return(sum(discrep^2))
}
err<-sse(theta,mf.params,dat)
#test
err
```




We could skip ahead to  optimisation here to simply find the "best" single parameter values given the model and the data. Or carry out more advance Bayeisan methods. Instead, to illustrate how this works and to examine the error surface, we will set up a simple grid of parameters. Because the models runs are not actually dependent on each other (they are sequential), we can also do this much more quickly with parallel computing. 

```{r}


# below not working



# two parameter grid
kappa <- seq(from=0.05,to=0.1,by=0.05)
effort <- seq(from=0,to=2,by=0.1)
grid <- expand.grid(kappa=kappa,effort=effort)
grid$SSE <- apply(grid,1,f)



#optimisation
vals<-optim(par=c(1e3,0.1),sse,method ="SANN")

# two parameter grid
kappa <- seq(from=0.05,to=0.1,by=0.05)
mort <- seq(from=0,to=2,by=0.1)
grid <- expand.grid(kappa=kappa,mort=mort)
grid$SSE <- apply(grid,1,sse)


```



###### Bayesian in a week


```{r}
## Step 1. generate parameter sets to sample and plug in to model
library(randtoolbox)
library(boot)
#### generate Round  1
theta <- sobol(n = 20000, dim = 26)
for(i in 1:12){
  # Rmax
  theta[,i] <- c(10,10,14,18,12,10,14,15)+ theta[,i] * c(20,19,16,13,18,20,16,15)
}
  #erepro
for(i in 13:24){
  theta[,i] <-  inv.logit(-5 + theta[,i] * 5)
}
 #kappa
theta[,25] <- 5 + theta[,25] * 25

 #background mort
theta[,26] <- inv.logit(-5 + theta[,26] * 5)


```



```{r}
#### Run Model (could set up on cluster, but nice to have prgress bar)

library(pbapply)
round_1 <-  pbapply(theta,1,runmodel, parms)
save(theta,round_1,file="data/round1.Rdata")

```

```{r}
# set up error function to compare predictions with observations (only need range of observations)
## Calculate the difference between the simulated catches and the observations. plot the results
load("data/round1.Rdata")
difs<- pbsapply(1:nrow(theta),function(x){colSums((matrix(round_1[,x],ncol=8) - log(dat[,-1]))^2,na.rm=T)})


```

```{r}
# build emulator on diffs
```


```{r}
# sample and cheick params work
```


```{r}
# MCMC - posterior
```


```{r}
# resample posterior and compare emulator wth mizer (get importance weightings)
```


```{r}
# resample posterior accordingto weights & run mize scenarios
```